---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# xpectr

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/xpectr)](https://CRAN.R-project.org/package=xpectr)
<!-- badges: end -->

`xpectr` provides a set of utilities and RStudio addins for generating tests for testthat unit testing.

**Author:** [Ludvig R. Olsen](http://ludvigolsen.dk/) ( r-pkgs@ludvigolsen.dk ) <br/>
**Started:** January 2020 

## Installation

You can install the development version with:

```{r eval=FALSE}
devtools::install_github("ludvigolsen/xpectr")
```


## Main functions 

### Generator functions

Note: `gxs` stands for `generate expectations`.

* `gxs_selection()` 
* `gxs_function()` 

### Helper functions
* `strip()`
* `strip_msg()`
* `smpl()`
* `element_lengths()`, `element_types()`, `element_classes()`
* `set_test_seed()`
* `prepare_insertion()`
* `capture_side_effects()`
* `capture_parse_eval_side_effects()`
* `stop_if()`, `warn_if()`, `message_if()`

## Addins

* `Insert Expectations` : generates `testthat` `expect_*` tests from selected code
* `dput() selected` : applies `dput()` to selected code
* `Insert checkmate AssertCollection code` : inserts code for initializing and reporting a checkmate AssertCollection
* `Wrap string with paste0` : splits selected string every n characters and wraps in `paste0` call

## Using in packages

It is suggested to add `xpectr` in the `Suggests` field in the `DESCRIPTION` file. 
This allows you to use it in unit tests and users aren't required to have it installed.

NOTE: This won't work on CRAN before it is released there.

## Examples

```{r warning=FALSE, message=FALSE}
library(xpectr)
library(testthat)
library(dplyr)

# Set a seed
# When R > 3.6.0, it sets sampling.kind to "Rounding" to make
# tests compatible with previous versions of R
set_test_seed(42)
```

```{r}
# Some data

num_vec <- 1:10
char_vec <- head(LETTERS, 10)
long_vec <- 1:40
a_factor <- factor(c("a","b","c"))

df <- data.frame(
  'a' = c(1, 2, 3),
  'b' = c('t', 'y', 'u'),
  "c" = a_factor,
  stringsAsFactors = FALSE
) %>% 
  dplyr::group_by(a)

# A function with side effects
fn <- function(raise = FALSE){
  message("Hi! I'm Kevin, your favorite message!")
  warning("G'Day Mam! I'm a warning to the world!")
  message("Kevin is ma name! Yesss!")
  warning("Hopefully the whole world will see me :o")
  if (isTRUE(raise)){
    stop("Lord Evil Error has arrived! Yeehaaa")
  }
}
```

### `gxs_selection()`

Note: `gxs_selection()` can be used with the `Insert Expectations` addin. See `?insertExpectationsAddin` for instructions on how to set up a key command.

#### Selection is a vector

Numeric vector:

```{r}
# Inspect num_vec
num_vec
```

```{r eval=FALSE}
# Generate expectations
gxs_selection("num_vec")

# Inserts the following tests:

## Testing 'num_vec'                                                        ####
## Initially generated by xpectr
# Testing class
expect_equal(
  class(num_vec),
  "integer",
  fixed = TRUE)
# Testing type
expect_type(
  num_vec,
  type = "integer")
# Testing values
expect_equal(
  num_vec,
  1:10,
  tolerance = 1e-4)
# Testing names
expect_equal(
  names(num_vec),
  NULL,
  fixed = TRUE)
# Testing length
expect_equal(
  length(num_vec),
  10L)
# Testing sum of element lengths
expect_equal(
  sum(xpectr::element_lengths(num_vec)),
  10L)
## Finished testing 'num_vec'                                               ####

```

Character vector:

```{r}
# Inspect char_vec
char_vec
```

```{r eval=FALSE}
# Generate expectations
gxs_selection("char_vec")

# Inserts the following tests:

## Testing 'char_vec'                                                       ####
## Initially generated by xpectr
# Testing class
expect_equal(
  class(char_vec),
  "character",
  fixed = TRUE)
# Testing type
expect_type(
  char_vec,
  type = "character")
# Testing values
expect_equal(
  char_vec,
  c("A", "B", "C", "D", "E", "F", "G", "H", "I", "J"),
  fixed = TRUE)
# Testing names
expect_equal(
  names(char_vec),
  NULL,
  fixed = TRUE)
# Testing length
expect_equal(
  length(char_vec),
  10L)
# Testing sum of element lengths
expect_equal(
  sum(xpectr::element_lengths(char_vec)),
  10L)
## Finished testing 'char_vec'                                              ####

```

Factor:

```{r}
# Inspect a_factor
a_factor
```

```{r eval=FALSE}
# Generate expectations
gxs_selection("a_factor")

# Inserts the following tests:

## Testing 'a_factor'                                                       ####
## Initially generated by xpectr
# Testing is factor
expect_true(
  is.factor(a_factor))
# Testing values
expect_equal(
  as.character(a_factor),
  c("a", "b", "c"),
  fixed = TRUE)
# Testing names
expect_equal(
  names(a_factor),
  NULL,
  fixed = TRUE)
# Testing length
expect_equal(
  length(a_factor),
  3L)
# Testing number of levels
expect_equal(
  nlevels(a_factor),
  3L)
# Testing levels
expect_equal(
  levels(a_factor),
  c("a", "b", "c"),
  fixed = TRUE)
## Finished testing 'a_factor'                                              ####

```

Long vector (currently > 30 elements) adds `smpl()`:

```{r}
# Inspect long_vec
long_vec
```

```{r eval=FALSE}
# Generate expectations
gxs_selection("long_vec")

# Inserts the following tests:

## Testing 'long_vec'                                                       ####
## Initially generated by xpectr
# Testing class
expect_equal(
  class(long_vec),
  "integer",
  fixed = TRUE)
# Testing type
expect_type(
  long_vec,
  type = "integer")
# Testing values
expect_equal(
  xpectr::smpl(long_vec, n = 30),
  c(2L, 3L, 5L, 6L, 7L, 8L, 10L, 11L, 12L, 13L, 14L, 17L, 18L, 19L,
    21L, 22L, 23L, 24L, 25L, 26L, 27L, 29L, 30L, 31L, 32L, 35L,
    36L, 37L, 38L, 40L),
  tolerance = 1e-4)
# Testing names
expect_equal(
  names(xpectr::smpl(long_vec, n = 30)),
  NULL,
  fixed = TRUE)
# Testing length
expect_equal(
  length(long_vec),
  40L)
# Testing sum of element lengths
expect_equal(
  sum(xpectr::element_lengths(long_vec)),
  40L)
## Finished testing 'long_vec'                                              ####

```

#### Selection is a data frame

```{r}
# Inspect df
df
```

```{r eval=FALSE}
# Generate expectations
gxs_selection("df")

# Inserts the following tests:

## Testing 'df'                                                             ####
## Initially generated by xpectr
# Testing class
expect_equal(
  class(df),
  c("grouped_df", "tbl_df", "tbl", "data.frame"),
  fixed = TRUE)
# Testing column values
expect_equal(
  df[["a"]],
  c(1, 2, 3),
  tolerance = 1e-4)
expect_equal(
  df[["b"]],
  c("t", "y", "u"),
  fixed = TRUE)
expect_equal(
  df[["c"]],
  structure(1:3, .Label = c("a", "b", "c"), class = "factor"))
# Testing column names
expect_equal(
  names(df),
  c("a", "b", "c"),
  fixed = TRUE)
# Testing column classes
expect_equal(
  xpectr::element_classes(df),
  c("numeric", "character", "factor"),
  fixed = TRUE)
# Testing column types
expect_equal(
  xpectr::element_types(df),
  c("double", "character", "integer"),
  fixed = TRUE)
# Testing dimensions
expect_equal(
  dim(df),
  c(3L, 3L))
# Testing group keys
expect_equal(
  colnames(dplyr::group_keys(df)),
  "a",
  fixed = TRUE)
## Finished testing 'df'                                                    ####

```

#### Selection is a function with side effects

Note that the `strip_msg` and `strip` functions are wrapped around the side effect and the message. These remove non-alphanumeric symbols from the tested strings. When running `testthat` unit tests on different systems, they sometimes vary in the use of punctuation and newlines (\\n). These functions will help the `expect_*` function to ignore that. In case such differences are important to catch, you can set `strip = FALSE` in `gxs_selection()`.

```{r}
# Inspect fn
fn
```

```{r eval=FALSE}
# Generate expectations
gxs_selection("fn()")

# Inserts the following tests:

## Testing 'fn()'                                                           ####
## Initially generated by xpectr
# Testing side effects
expect_warning(
  xpectr::strip_msg(fn()),
  xpectr::strip("G'Day Mam! I'm a warning to the world!"),
  fixed = TRUE)
expect_warning(
  xpectr::strip_msg(fn()),
  xpectr::strip("Hopefully the whole world will see me :o"),
  fixed = TRUE)
expect_message(
  xpectr::strip_msg(fn()),
  xpectr::strip("Hi! I'm Kevin, your favorite message!\n"),
  fixed = TRUE)
expect_message(
  xpectr::strip_msg(fn()),
  xpectr::strip("Kevin is ma name! Yesss!\n"),
  fixed = TRUE)
## Finished testing 'fn()'                                                  ####

# In case of errors, the warnings and messages aren't tested

gxs_selection("fn(raise = TRUE)")

# Inserts the following tests:

## Testing 'fn(raise = TRUE)'                                               ####
## Initially generated by xpectr
# Testing side effects
expect_error(
  xpectr::strip_msg(fn(raise = TRUE)),
  xpectr::strip("Lord Evil Error has arrived! Yeehaaa"),
  fixed = TRUE)
## Finished testing 'fn(raise = TRUE)'                                      ####

```

### `gxs_function()`

Based on a set of supplied values for each function argument, a set of `testthat` `expect_*` statements are generated.

The first value supplied for an argument is considered the _valid baseline_ value. For each argument, we create tests for each of the supplied values, where the other arguments have their baseline value. It automatically adds `NULL` to the list of values to check for each argument.

```{r eval=FALSE}
# Define a function with arguments
fn <- function(x, y, z) {
  if (x > 3) stop("'x' > 3")
  if (y < 0) warning("'y'<0")
  if (z == 10) message("'z' was 10!")
  x + y + z
}

# Create tests for the function
# Note: We currently need to specify the list of arguments
# in the function call
gxs_function(fn = fn,
             args_values = list(
               "x" = list(2, 4, NA),
               "y" = list(0,-1),
               "z" = list(5, 10, 15)
             ))

# Inserts the following tests:

## Testing 'fn'                                                             ####
## Initially generated by xpectr
# Testing different combinations of argument values

# Testing fn(x = 4, y = 0, z = 5)
# Testing side effects
expect_error(
  xpectr::strip_msg(fn(x = 4, y = 0, z = 5)),
  xpectr::strip("'x' > 3"),
  fixed = TRUE)

# Testing fn(x = NA, y = 0, z = 5)
# Testing side effects
expect_error(
  xpectr::strip_msg(fn(x = NA, y = 0, z = 5)),
  xpectr::strip("missing value where TRUE/FALSE needed"),
  fixed = TRUE)

# Testing fn(x = 2, y = -1, z = 5)
# Testing side effects
expect_warning(
  xpectr::strip_msg(fn(x = 2, y = -1, z = 5)),
  xpectr::strip("'y'<0"),
  fixed = TRUE)

# Testing fn(x = 2, y = 0, z = 10)
# Testing side effects
expect_message(
  xpectr::strip_msg(fn(x = 2, y = 0, z = 10)),
  xpectr::strip("'z' was 10!\n"),
  fixed = TRUE)

# Testing fn(x = 2, y = 0, z = 15)
# Testing class
expect_equal(
  class(fn(x = 2, y = 0, z = 15)),
  "numeric",
  fixed = TRUE)
# Testing type
expect_type(
  fn(x = 2, y = 0, z = 15),
  type = "double")
# Testing values
expect_equal(
  fn(x = 2, y = 0, z = 15),
  17,
  tolerance = 1e-4)
# Testing names
expect_equal(
  names(fn(x = 2, y = 0, z = 15)),
  NULL,
  fixed = TRUE)
# Testing length
expect_equal(
  length(fn(x = 2, y = 0, z = 15)),
  1L)
# Testing sum of element lengths
expect_equal(
  sum(xpectr::element_lengths(fn(x = 2, y = 0, z = 15))),
  1L)

# Testing fn(x = 2, y = 0, z = 5)
# Testing class
expect_equal(
  class(fn(x = 2, y = 0, z = 5)),
  "numeric",
  fixed = TRUE)
# Testing type
expect_type(
  fn(x = 2, y = 0, z = 5),
  type = "double")
# Testing values
expect_equal(
  fn(x = 2, y = 0, z = 5),
  7,
  tolerance = 1e-4)
# Testing names
expect_equal(
  names(fn(x = 2, y = 0, z = 5)),
  NULL,
  fixed = TRUE)
# Testing length
expect_equal(
  length(fn(x = 2, y = 0, z = 5)),
  1L)
# Testing sum of element lengths
expect_equal(
  sum(xpectr::element_lengths(fn(x = 2, y = 0, z = 5))),
  1L)

# Testing fn(x = NULL, y = 0, z = 5)
# Testing side effects
expect_error(
  xpectr::strip_msg(fn(x = NULL, y = 0, z = 5)),
  xpectr::strip("argument is of length zero"),
  fixed = TRUE)

# Testing fn(x = 2, y = NULL, z = 5)
# Testing side effects
expect_error(
  xpectr::strip_msg(fn(x = 2, y = NULL, z = 5)),
  xpectr::strip("argument is of length zero"),
  fixed = TRUE)

# Testing fn(x = 2, y = 0, z = NULL)
# Testing side effects
expect_error(
  xpectr::strip_msg(fn(x = 2, y = 0, z = NULL)),
  xpectr::strip("argument is of length zero"),
  fixed = TRUE)

## Finished testing 'fn'                                                    ####

```


### `wrapStringAddin()`

This addin splits long strings and wraps them with `paste0()`.

```{r eval=FALSE}
wrapStringAddin("This is a fairly long sentence that we would very very much like to make shorter in our test file!")

# Inserts the following:

paste0("This is a fairly long sentence that we would very very much ",
       "like to make shorter in our test file!")

```

