---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# xpectr

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/xpectr)](https://CRAN.R-project.org/package=xpectr)
<!-- badges: end -->

`xpectr` provides a set of utilities and RStudio addins for generating tests for testthat unittesting.

**Author:** [Ludvig R. Olsen](http://ludvigolsen.dk/) ( r-pkgs@ludvigolsen.dk ) <br/>
**Started:** January 2020 

## Installation

You can install the development version with:

```{r eval=FALSE}
devtools::install_github("ludvigolsen/xpectr")
```


## Main functions

Note: `gxs` stands for `generate expectations`.

* `gxs_selection()` 
* `strip()`
* `strip_msg()`
* `smpl()`
* `set_seed()`
* `prepare_insertion()`
* `capture_side_effects()`
* `capture_parse_eval_side_effects()`
* `stop_if()`, `warn_if()`, `message_if()`

## Addins

* `Insert Expectations` : generates `testthat` `expect_*` tests from selected code
* `dput() selected` : applies `dput()` to selected code
* `Insert checkmate AssertCollection code` : inserts code for initializing and reporting a checkmate AssertCollection
* `Wrap string with paste0` : splits selected string every n characters and wraps in `paste0` call

## Using in packages

It is suggested to add `xpectr` in the `Suggests` field in the `DESCRIPTION` file. 
This allows you to use it in unit tests and users aren't required to have it installed.

NOTE: This won't work on CRAN before it is released there.

## Examples

```{r}
library(xpectr)
library(testthat)

# Set a seed
# When R > 3.6.0 it sets sampling.kind to "Rounding" to make
# tests compatible with previous versions of R
set_seed(42)
```

```{r}
# Some data

num_vec <- 1:10
char_vec <- head(LETTERS, 10)
long_vec <- 1:40

df <- data.frame(
  'a' = c(1, 2, 3),
  'b' = c('t', 'y', 'u'),
  stringsAsFactors = FALSE
)

# A function with side effects
fn <- function(raise = FALSE){
  message("Hi! I'm Kevin, your favorite message!")
  warning("G'Day Mam! I'm a warning to the world!")
  message("Kevin is ma name! Yesss!")
  warning("Hopefully the whole world will see me :o")
  if (isTRUE(raise)){
    stop("Lord Evil Error has arrived! Yeehaaa")
  }
}
```

### `gxs_selection()`

Note: `gxs_selection()` can be used with the `Insert Expectations` addin. See `?insertExpectationsAddin` for instructions on how to set up a key command.

#### Selection is a vector

Numeric vector:
```{r eval=FALSE}
gxs_selection("num_vec")

# Inserts the following tests:

expect_equal(
  num_vec,
  1:10,
  tolerance = 1e-4)
```

Character vector:
```{r eval=FALSE}
gxs_selection("char_vec")

# Inserts the following tests:

expect_equal(
  char_vec,
  c("A", "B", "C", "D", "E", "F", "G", "H", "I", "J"),
  fixed = TRUE)
```

Long vector (currently > 30 elements) adds `smpl()`:
```{r eval=FALSE}
gxs_selection("long_vec")

# Inserts the following tests:

expect_equal(
  xpectr::smpl(long_vec, n = 30),
  c(2L, 3L, 5L, 6L, 7L, 8L, 10L, 11L, 12L, 13L, 14L, 17L, 18L, 19L, 21L, 22L, 23L, 24L, 25L, 26L, 27L, 29L, 30L, 31L, 32L, 35L, 36L, 37L, 38L, 40L),
  tolerance = 1e-4)
```

#### Selection is a data frame

```{r eval=FALSE}
gxs_selection("df")

# Inserts the following tests:

expect_equal(
  df[["a"]],
  c(1, 2, 3),
  tolerance = 1e-4)
expect_equal(
  df[["b"]],
  c("t", "y", "u"),
  fixed = TRUE)
expect_equal(
  names( df ),
  c("a", "b"),
  fixed = TRUE)
```

#### Selection is a function with side effects

Note that the `strip_msg` and `strip` functions are wrapped around the side effect and the message. These remove non-alphanumeric symbols from the tested strings. When running `testthat` unit tests on different systems, they sometimes vary in the use of punctuation and newlines (\\n). These functions will help the `expect_*` function to ignore that. In case such differences are important to catch, you can set `strip = FALSE` in `gxs_selection()`.

```{r eval = FALSE}
gxs_selection("fn()")

# Inserts the following tests:

expect_warning(
  xpectr::strip_msg(fn()),
  xpectr::strip("G'Day Mam! I'm a warning to the world!"),
  fixed = TRUE)
expect_warning(
  xpectr::strip_msg(fn()),
  xpectr::strip("Hopefully the whole world will see me :o"),
  fixed = TRUE)
expect_message(
  xpectr::strip_msg(fn()),
  xpectr::strip("Hi! I'm Kevin, your favorite message!\n"),
  fixed = TRUE)
expect_message(
  xpectr::strip_msg(fn()),
  xpectr::strip("Kevin is ma name! Yesss!\n"),
  fixed = TRUE)

# In case of errors, the warnings and messages aren't tested

gxs_selection("fn(raise = TRUE)")

# Inserts the following tests:

expect_error(
  xpectr::strip_msg(fn(raise = TRUE)),
  xpectr::strip("Lord Evil Error has arrived! Yeehaaa"),
  fixed = TRUE)
```

### wrapStringAddin()

This addin splits long strings and wraps them with `paste0()`.

```{r}
wrapStringAddin("This is a fairly long sentence that we would very very much like to make shorter in our test file!")

# Inserts the following:

paste0("This is a fairly long sentence that we would very very much ",
       "like to make shorter in our test file!")```

