---
title: "README for xpectr"
author: 
  - "Ludvig Renbo Olsen"
date: "`r Sys.Date()`"
abstract: |
  This vignette is an introduction to the `xpectr` package. 
  `xpectr` provides a set of utilities and RStudio addins for generating tests for 
  `testthat` unit testing.
  &nbsp;  
  &nbsp;  
  Contact author at r-pkgs@ludvigolsen.dk
  &nbsp;  
  &nbsp;  
  
output: 
  rmarkdown::html_vignette:
    css: 
    - !expr system.file("rmarkdown/templates/html_vignette/resources/vignette.css", package = "rmarkdown")
    - styles.css
    fig_width: 6
    fig_height: 4
    toc: yes
    number_sections: no
  rmarkdown::pdf_document:
    highlight: tango
    number_sections: yes
    toc: yes
    toc_depth: 4
vignette: >
  %\VignetteIndexEntry{Introduction_to_cvms}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/vignette_intro-",
  dpi = 92,
  fig.retina = 2
)
```


When developing R packages, it's good practice to build a good set of unit tests that will notify you when something breaks in the future. For this, the [`testthat`](https://testthat.r-lib.org/) package is commonly used. 
Often though, we end up writing a similar set of tests again and again, which can be both tedious and time-consuming. 

`xpectr` helps you by generating common `testthat` tests. Its goal is **not** to replace the active thinking process of finding meaningful tests for your functions, but to help **systematize** and **ease** that process. 

One such example is `gxs_function()`, which generates tests for a set of argument value combinations. This allows you to focus on the selection of argument values to test. Once the tests have been generated, you can go through each of them to ensure your function is working as intended and to add relevant tests. 

Note: If you comment out the call to `gxs_function()`, it is easy to later regenerate the tests. By using a diff tool, you can check that only the intended changes have been made to the file.

## Installation

You can install the development version with:

```{r eval=FALSE}
# install.packages("devtools")  
devtools::install_github("ludvigolsen/xpectr")
```

You can install the latest CRAN release of the package with:

```{r eval=FALSE}
install.packages("xpectr")  
```

## Main functions 

### Generator functions

These functions are used for *generating expectations* (gxs). 

* `gxs_selection()` 
* `gxs_function()` 

### Functions for use in tests

* `strip()`
* `strip_msg()`
* `suppress_mw()`
* `capture_side_effects()`
* `smpl()`
* `simplified_formals()`
* `element_lengths()`, `element_types()`, `element_classes()`
* `num_total_elements()`
* `set_test_seed()`

### Helper functions

* `prepare_insertion()`
* `capture_parse_eval_side_effects()`
* `stop_if()`, `warn_if()`, `message_if()`

## Addins

* `Insert Expectations` : generates `testthat` `expect_*` tests from selected code (with `gxs_selection()`)
* `dput() selected` : applies `dput()` to selected code
* `Wrap string with paste0` : splits selected string every n characters and wraps in `paste0` call
* `Insert checkmate AssertCollection code` : inserts code for initializing and reporting a checkmate AssertCollection

## Using in packages

Suggestion: Add `xpectr` in the `Suggests` field in the `DESCRIPTION` file.

## Examples

```{r warning=FALSE, message=FALSE}
library(xpectr)
library(testthat)
library(dplyr)

# Set a seed
# When R > 3.6.0, it sets sampling.kind to "Rounding" to make
# tests compatible with previous versions of R
set_test_seed(42)
```

```{r}
# Some data
num_vec <- 1:10
long_vec <- c(LETTERS, letters)
a_factor <- factor(c("a","b","c"))

df <- data.frame(
  'a' = c(1, 2, 3),
  'b' = c('t', 'y', 'u'),
  "c" = a_factor,
  stringsAsFactors = FALSE
) %>% 
  dplyr::group_by(a)

# A function with side effects
fn <- function(raise = FALSE){
  message("Hi! I'm Kevin, your favorite message!")
  warning("G'Day Mam! I'm a warning to the world!")
  message("Kevin is ma name! Yesss!")
  warning("Hopefully the whole world will see me :o")
  if (isTRUE(raise)){
    stop("Lord Evil Error has arrived! Yeehaaa")
  }
  "the output"
}
```

### gxs_selection

Note: `gxs_selection()` can be used with the `Insert Expectations` addin. See `?insertExpectationsAddin` for instructions on how to set up a key command.

#### Selection is a vector

##### Numeric vector

```{r}
# Inspect num_vec
num_vec
```

```{r eval=FALSE}
# Generate expectations
gxs_selection("num_vec")

# Inserts the following tests:

## Testing 'num_vec'                                                        ####
## Initially generated by xpectr
xpectr::set_test_seed(42)
# Testing class
expect_equal(
  class(num_vec),
  "integer",
  fixed = TRUE)
# Testing type
expect_type(
  num_vec,
  type = "integer")
# Testing values
expect_equal(
  num_vec,
  c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10),
  tolerance = 1e-4)
# Testing names
expect_equal(
  names(num_vec),
  NULL,
  fixed = TRUE)
# Testing length
expect_equal(
  length(num_vec),
  10L)
# Testing sum of element lengths
expect_equal(
  sum(xpectr::element_lengths(num_vec)),
  10L)
## Finished testing 'num_vec'                                               ####

```

##### Factor

```{r}
# Inspect a_factor
a_factor
```

```{r eval=FALSE}
# Generate expectations
gxs_selection("a_factor")

# Inserts the following tests:

## Testing 'a_factor'                                                       ####
## Initially generated by xpectr
xpectr::set_test_seed(42)
# Testing is factor
expect_true(
  is.factor(a_factor))
# Testing values
expect_equal(
  as.character(a_factor),
  c("a", "b", "c"),
  fixed = TRUE)
# Testing names
expect_equal(
  names(a_factor),
  NULL,
  fixed = TRUE)
# Testing length
expect_equal(
  length(a_factor),
  3L)
# Testing number of levels
expect_equal(
  nlevels(a_factor),
  3L)
# Testing levels
expect_equal(
  levels(a_factor),
  c("a", "b", "c"),
  fixed = TRUE)
## Finished testing 'a_factor'                                              ####

```

##### Long vector (sampling)

By default, vectors with more than 30 elements will be sampled. This adds `smpl()`, which temporarily sets a seed to make sure the same elements are returned every time.

```{r}
# Inspect long_vec
long_vec
```

```{r eval=FALSE}
# Generate expectations
gxs_selection("long_vec")

# Inserts the following tests:

## Testing 'long_vec'                                                       ####
## Initially generated by xpectr
xpectr::set_test_seed(42)
# Testing class
expect_equal(
  class(long_vec),
  "character",
  fixed = TRUE)
# Testing type
expect_type(
  long_vec,
  type = "character")
# Testing values
expect_equal(
  xpectr::smpl(long_vec, n = 30),
  c("C", "E", "G", "J", "K", "N", "O", "Q", "R", "S", "T", "W", "Y", 
    "Z", "b", "c", "d", "e", "h", "i", "j", "k", "l", "o", "p", 
    "r", "v", "w", "y", "z"),
  fixed = TRUE)
# Testing names
expect_equal(
  names(xpectr::smpl(long_vec, n = 30)),
  NULL,
  fixed = TRUE)
# Testing length
expect_equal(
  length(long_vec),
  52L)
# Testing sum of element lengths
expect_equal(
  sum(xpectr::element_lengths(long_vec)),
  52L)
## Finished testing 'long_vec'                                              ####

```

#### Selection is a data frame

Data frames are tested columnwise.

```{r}
# Inspect df
df
```

```{r eval=FALSE}
# Generate expectations
gxs_selection("df")

# Inserts the following tests:

## Testing 'df'                                                             ####
## Initially generated by xpectr
xpectr::set_test_seed(42)
# Testing class
expect_equal(
  class(df),
  c("grouped_df", "tbl_df", "tbl", "data.frame"),
  fixed = TRUE)
# Testing column values
expect_equal(
  df[["a"]],
  c(1, 2, 3),
  tolerance = 1e-4)
expect_equal(
  df[["b"]],
  c("t", "y", "u"),
  fixed = TRUE)
expect_equal(
  df[["c"]],
  structure(1:3, .Label = c("a", "b", "c"), class = "factor"))
# Testing column names
expect_equal(
  names(df),
  c("a", "b", "c"),
  fixed = TRUE)
# Testing column classes
expect_equal(
  xpectr::element_classes(df),
  c("numeric", "character", "factor"),
  fixed = TRUE)
# Testing column types
expect_equal(
  xpectr::element_types(df),
  c("double", "character", "integer"),
  fixed = TRUE)
# Testing dimensions
expect_equal(
  dim(df),
  c(3L, 3L))
# Testing group keys
expect_equal(
  colnames(dplyr::group_keys(df)),
  "a",
  fixed = TRUE)
## Finished testing 'df'                                                    ####

```

#### Selection is a function call with side effects

When the selected code generates an error, warning or message, we can test those as well. An error is tested with `expect_error()`, while messages and warnings are tested as character vectors, to make sure we catch additional warnings/messages. 

When running `testthat` unit tests on different systems, they sometimes vary in the use of punctuation and newlines (\\n). The `strip()` and `strip_msg()` functions are wrapped around the side effects to remove non-alphanumeric symbols from the tested strings. This helps the `expect_*` functions to ignore those differences. In cases where such differences are important to catch, you can set `strip = FALSE` in `gxs_selection()`.

We assign the output of the function to an `output_12345` variable, so we don't have to run it more than once. The number is randomly generated and is **not** guaranteed to be unique. `suppress_mw()` suppresses the messages and warnings.

```{r}
# Inspect fn
fn
```

```{r eval=FALSE}
# Generate expectations
gxs_selection("fn()")

# Inserts the following tests:

## Testing 'fn()'                                                           ####
## Initially generated by xpectr
xpectr::set_test_seed(42)
# Testing side effects
# Assigning side effects
side_effects_19148 <- xpectr::capture_side_effects(fn(), reset_seed = TRUE)
expect_equal(
  xpectr::strip(side_effects_19148[['warnings']]),
  xpectr::strip(c("G'Day Mam! I'm a warning to the world!", 
                  "Hopefully the whole world will see me :o")),
  fixed = TRUE)
expect_equal(
  xpectr::strip(side_effects_19148[['messages']]),
  xpectr::strip(c("Hi! I'm Kevin, your favorite message!\n", 
                  "Kevin is ma name! Yesss!\n")),
  fixed = TRUE)
# Assigning output
output_19148 <- xpectr::suppress_mw(fn())
# Testing class
expect_equal(
  class(output_19148),
  "character",
  fixed = TRUE)
# Testing type
expect_type(
  output_19148,
  type = "character")
# Testing values
expect_equal(
  output_19148,
  "the output",
  fixed = TRUE)
# Testing names
expect_equal(
  names(output_19148),
  NULL,
  fixed = TRUE)
# Testing length
expect_equal(
  length(output_19148),
  1L)
# Testing sum of element lengths
expect_equal(
  sum(xpectr::element_lengths(output_19148)),
  1L)
## Finished testing 'fn()'                                                  ####

# In case of errors, the warnings and messages aren't tested

gxs_selection("fn(raise = TRUE)")

# Inserts the following tests:

## Testing 'fn(raise = TRUE)'                                               ####
## Initially generated by xpectr
xpectr::set_test_seed(42)
# Testing side effects
expect_error(
  xpectr::strip_msg(fn(raise = TRUE)),
  xpectr::strip("Lord Evil Error has arrived! Yeehaaa"),
  fixed = TRUE)
## Finished testing 'fn(raise = TRUE)'                                      ####

```

### gxs_function

When testing the inputs to a function, `gxs_function()` allows us to quickly specify values to check and generates tests for each of them.

The first value supplied for an argument is considered the *valid baseline* value. For each argument, we create tests for each of the supplied values, where the other arguments have their baseline value. 

By default, each argument is tested with the `NULL` object as well, why we only need to specify it when the baseline value should be `NULL`.

It is important that we manually read through the generated tests to make sure that our function is behaving as intended, and to check if any important tests are missing.

```{r eval=FALSE}
# Define a function with arguments
fn <- function(x, y, z) {
  if (x > 3) stop("'x' > 3")
  if (y < 0) warning("'y'<0")
  if (z == 10) message("'z' was 10!")
  x + y + z
}

# Create tests for the function
# Note: We currently need to specify the list of arguments
# in the function call
gxs_function(fn = fn,
             args_values = list(
               "x" = list(2, 4, NA),
               "y" = list(0,-1),
               "z" = list(5, 10)
             ))

# Inserts the following tests:

## Testing 'fn'                                                             ####
## Initially generated by xpectr
# Testing different combinations of argument values

# Testing fn(x = 2, y = 0, z = 5)
xpectr::set_test_seed(42)
# Assigning output
output_19148 <- fn(x = 2, y = 0, z = 5)
# Testing class
expect_equal(
  class(output_19148),
  "numeric",
  fixed = TRUE)
# Testing type
expect_type(
  output_19148,
  type = "double")
# Testing values
expect_equal(
  output_19148,
  7,
  tolerance = 1e-4)
# Testing names
expect_equal(
  names(output_19148),
  NULL,
  fixed = TRUE)
# Testing length
expect_equal(
  length(output_19148),
  1L)
# Testing sum of element lengths
expect_equal(
  sum(xpectr::element_lengths(output_19148)),
  1L)

# Testing fn(x = 4, y = 0, z = 5)
# Changed from baseline: x
xpectr::set_test_seed(42)
# Testing side effects
expect_error(
  xpectr::strip_msg(fn(x = 4, y = 0, z = 5)),
  xpectr::strip("'x' > 3"),
  fixed = TRUE)

# Testing fn(x = NA, y = 0, z = 5)
# Changed from baseline: x
xpectr::set_test_seed(42)
# Testing side effects
expect_error(
  xpectr::strip_msg(fn(x = NA, y = 0, z = 5)),
  xpectr::strip("missing value where TRUE/FALSE needed"),
  fixed = TRUE)

# Testing fn(x = NULL, y = 0, z = 5)
# Changed from baseline: x
xpectr::set_test_seed(42)
# Testing side effects
expect_error(
  xpectr::strip_msg(fn(x = NULL, y = 0, z = 5)),
  xpectr::strip("argument is of length zero"),
  fixed = TRUE)

# Testing fn(x = 2, y = -1, z = 5)
# Changed from baseline: y
xpectr::set_test_seed(42)
# Testing side effects
# Assigning side effects
side_effects_16417 <- xpectr::capture_side_effects(fn(x = 2, y = -1, z = 5), reset_seed = TRUE)
expect_equal(
  xpectr::strip(side_effects_16417[['warnings']]),
  xpectr::strip("'y'<0"),
  fixed = TRUE)
expect_equal(
  xpectr::strip(side_effects_16417[['messages']]),
  xpectr::strip(character(0)),
  fixed = TRUE)
# Assigning output
output_16417 <- xpectr::suppress_mw(fn(x = 2, y = -1, z = 5))
# Testing class
expect_equal(
  class(output_16417),
  "numeric",
  fixed = TRUE)
# Testing type
expect_type(
  output_16417,
  type = "double")
# Testing values
expect_equal(
  output_16417,
  6,
  tolerance = 1e-4)
# Testing names
expect_equal(
  names(output_16417),
  NULL,
  fixed = TRUE)
# Testing length
expect_equal(
  length(output_16417),
  1L)
# Testing sum of element lengths
expect_equal(
  sum(xpectr::element_lengths(output_16417)),
  1L)

# Testing fn(x = 2, y = NULL, z = 5)
# Changed from baseline: y
xpectr::set_test_seed(42)
# Testing side effects
expect_error(
  xpectr::strip_msg(fn(x = 2, y = NULL, z = 5)),
  xpectr::strip("argument is of length zero"),
  fixed = TRUE)

# Testing fn(x = 2, y = 0, z = 10)
# Changed from baseline: z
xpectr::set_test_seed(42)
# Testing side effects
# Assigning side effects
side_effects_17365 <- xpectr::capture_side_effects(fn(x = 2, y = 0, z = 10), reset_seed = TRUE)
expect_equal(
  xpectr::strip(side_effects_17365[['warnings']]),
  xpectr::strip(character(0)),
  fixed = TRUE)
expect_equal(
  xpectr::strip(side_effects_17365[['messages']]),
  xpectr::strip("'z' was 10!\n"),
  fixed = TRUE)
# Assigning output
output_17365 <- xpectr::suppress_mw(fn(x = 2, y = 0, z = 10))
# Testing class
expect_equal(
  class(output_17365),
  "numeric",
  fixed = TRUE)
# Testing type
expect_type(
  output_17365,
  type = "double")
# Testing values
expect_equal(
  output_17365,
  12,
  tolerance = 1e-4)
# Testing names
expect_equal(
  names(output_17365),
  NULL,
  fixed = TRUE)
# Testing length
expect_equal(
  length(output_17365),
  1L)
# Testing sum of element lengths
expect_equal(
  sum(xpectr::element_lengths(output_17365)),
  1L)

# Testing fn(x = 2, y = 0, z = NULL)
# Changed from baseline: z
xpectr::set_test_seed(42)
# Testing side effects
expect_error(
  xpectr::strip_msg(fn(x = 2, y = 0, z = NULL)),
  xpectr::strip("argument is of length zero"),
  fixed = TRUE)

## Finished testing 'fn'                                                    ####

```


### wrapStringAddin

The `wrapStringAddin` splits long strings and wraps them with `paste0()`.

```{r eval=FALSE}
wrapStringAddin("This is a fairly long sentence that we would very very much like to make shorter in our test file!")

# Inserts the following:

paste0("This is a fairly long sentence that we would very very much ",
       "like to make shorter in our test file!")

```

